Practical Optional Types for Clojure

### 2015-02-25 formal-model.tex
hmap examples
review first half of formal model

We describe our type system in a similar way to Typed Racket (\citet{TF10}),
with differences highlighted in $\trdiff{\text{blue}}$.
The typing judgement 
$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}.

@@

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.

Values include booleans, \nil{}, class literals, keywords (written with a : prefix like \makekw{a}), 
constants and closures.  \false{} and \nil{} are logically false; all other values are logically true.

@@

Propositions are identical to~\cite{TF10}---positive and negative type propositions, implications,
conjunctions, disjuctions, trivial and impossible propositions.
Type propositions contain a non-empty object which it claims type information about.

Proposition environments are sequences of propositions in conjunctive normal form.

Figure ~\ref{main:figure:typingrules} contains the standard typing rules.
The operational semantics are standard, except \nil{} is logically false:

@@

    %{\TDo}
    %{\TClass}
    {\TIf}
    {\TLet}

@@

\subsection{Reasoning about Exceptional Control Flow}

Clojure assertions expand to simple exceptional control flow.

\begin{minted}{clojure}
(fn [x :- (U nil Number)] 
  (assert (number? x))
  (inc x))
\end{minted}

roughly expands to

\begin{minted}{clojure}
(fn [x :- (U nil Number)]
  (do (if (number? x) nil (throw (new Exception)))
      (inc x)))
\end{minted}

@@

\smallskip

B-Do simply evaluates its arguments sequentially and returns the right argument.
Since errors are not values, we define error propagation semantics
like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).

\begin{mathpar}
    {\BDo}

\infer [BE-Error]
{}
{ \opsem {\openv{}} 
         {\errorvalv{}}
         {\errorvalv{}}}

\infer [BE-Do1]
{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
\end{mathpar}

@@

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \clj{(inc :a)} is unreachable (and it clearly is by BE-Do1).

This works similarly with our motivating example. Intuitively, when evaluating
\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
so we assume \clj{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \clj{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \clj{x} as a \Number.

\subsection{Heterogenous maps}

@@

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
\clj{(HMap :mandatory p :absent-keys a)} is written \HMapgeneric{\mandatory{}}{\absent{}}.
The metavariable \completenessmeta{} tracks whether a HMap is fully ({\complete{}}) or partially ({\partial{}})
specified. \mandatory{} specifies a set of known present keyword entries, \absent{} tracks known absent keyword
entries and we abbreviate \HMapcwithabsent{\mandatory{}}{\absent{}} as \HMapc{\mandatory{}}. Keys cannot be both present and absent.

\getliteral{} and \assocliteral{} lookup and extend immutable maps respectively.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

We introduce our first path element {\keype{\k{}}}, the symbolic representation
of a \getliteral{} expression on keyword \k{} where the sentinel value for
a missing entry is \nil{}.

Figure~\ref{main:figure:hmapsem} shows the semantics for HMaps. B-Get returns the associated
value if the entry is present otherwise  B-GetMissing returns \nil{}. B-Assoc returns a new extended
map.

Typing rules are given in figure~\ref{main:figure:hmaptyping}. T-GetHMap handles the simple
case where we know a lookup will succeed and use B-Get. In T-GetHMapAbsent we know B-GetMissing will give
a \Nil{} return type.
T-GetHMapPartialDefault does not know if the key is present or absent and has a \Top{} return type.
Even though their types vary significantly, we still use the same object.
T-Assoc extends the known keyword entries, overwriting previous entries with the same key.

@@

The \classconst{} function is modeled as a constant which returns
an object's \Class{}, or \nil{} if provided \nil{}.

@@

Our second path element \classpe{} is the symbolic representation
of a call to \classconst{}.

$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

% TODO semantics

@@

\begin{mathpar}
  {\LUpdate}
\end{mathpar}

For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
and
\inpropenv {\propenv{}} {\notprop {\Nil{}} {\x{}}},
then
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype{\Nil{}}} {\emptypath{}}} {\x{}}},
$$
so we learn \inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).

Now that we have our path elements \classpe{} and \keype{\k{}}
and the typings rules that introduce them, we look at \updateliteral{} 
(figure~\ref{main:figure:update}) and give one interesting example for each.

@@

\begin{minted}{clojure}
(fn [x :- (U nil Keyword)]
  (when (class x)
    ; here
    ))
\end{minted}

@@

From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
The test has the object {\path{\classpe{}}{\x{}}} by T-App.
By T-If we can assume
{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
{\path{\classpe{}}{\x{}}}, 
by T-Update 
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
$$

This hits the sixth \classpe{} case of \updateliteral{}, which returns
$$
{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
$$
deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).

@@

\begin{minted}{clojure}
(fn [x :- '{:kw (U nil Keyword)}]
  (when (get x :kw)
    ; here
    ))
\end{minted}

@@

$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$

rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{}.

$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$

infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.

@@

We can use \isaliteral{} as a crude \nil{} test

\inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}

@@

Substitution is used to ensure well formed propositions.

The final case gives up; we cannot distinguish between equality and subclassing mode.

@@

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. The above is now written

%$\letexp{foo}{\defmultiexp {\ArrowOne{\x{}}{\Top}{\clj{String}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}}
%                       {\classconst{}}}
%                     {\\ bar}$

\begin{minted}{clojure}
(let [foo (defmulti [Any -> String] class)
      foo (defmethod foo Number [x] (str (inc x)))
      foo (defmethod foo Keyword [x] (str (name x)))]
  (foo :a)) ;=> "a"
\end{minted}

@@

{\footnotesize
\constantsemfigure{}

{\convertjavatypefigure{figure}{\footnotesize}}
}

>>>>

We formalise our type system like \citet{TF10}
with differences highlighted in $\trdiff{\text{blue}}$.
The typing judgement 
$$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}. We sometimes write 
{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.

@@

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
constants and closures. \false{} and \nil{} are logically false, otherwise values are true.

@@

Propositions are standard---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
say object {\path {\pathelem{}} {\x{}}} is and is not of type \t{}
respectively.
We include implications,
conjunctions, disjuctions, and the trivial and impossible propositions.
Proposition environments are sequences of propositions.

Figure~\ref{main:figure:typingrules} contains the standard typing rules.
The operational semantics are standard, except \nil{} is logically false:

@@

    %{\TDo}
    %{\TClass}
    {\TIf}
    {\TLet}
    {\TConst}

@@

\subsection{Reasoning about Exceptional Control Flow}
\label{sec:doformal}

@@

\smallskip
%
%B-Do simply evaluates its arguments sequentially and returns the right argument.
%Since errors are not values, we define error propagation semantics
%like BE-Do1 (figure~\ref{appendix:figure:errorstuck} for the full rules).
%
%\begin{mathpar}
%    {\BDo}
%
%\infer [BE-Error]
%{}
%{ \opsem {\openv{}} 
%         {\errorvalv{}}
%         {\errorvalv{}}}
%
%\infer [BE-Do1]
%{ \opsem {\openv{}} {\e{1}} {\errorvalv{}} }
%{ \opsem {\openv{}} {\doexp{\e{1}}{\e{}}} {\errorvalv{}}}
%\end{mathpar}

@@

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \clj{(inc :a)} is unreachable.

Now we can check example~\ref{example:doexception}.
Intuitively, when evaluating
\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
so we assume \clj{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \clj{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \clj{x} as a \Number.

\subsection{Heterogenous maps}
\label{sec:hmapformal}

@@

We now present \HMapliteral{} types in detail.
Figure~\ref{main:figure:hmapsyntax} extends the syntax for terms and types.
$$
\clj{(HMap :mandatory p :absent-keys a :complete? b)}
$$
is written \HMapgeneric{\mandatory{}}{\absent{}},
where the metavariable \completenessmeta{} is {\complete{}} if \clj{b} is \clj{true} otherwise {\partial{}}. 
\mandatory{} specifies a set of known present keyword entries. \absent{} tracks known absent keys.
\HMapcwithabsent{\mandatory{}}{\absent{}} is abbreviated to \HMapc{\mandatory{}}. Keys cannot be both present and absent.

\clj{(get m :a)} is identical to \clj{(:a m)}.
To simplify presentation, we only provide syntax for the empty map literal and
restrict lookup and extension to keyword keys. The metavariable \mapval{}
ranges over the runtime value of maps {\curlymapvaloverright{\k{}}{\v{}}}.

We introduce our first path element {\keype{\k{}}}, the symbolic representation
of \clj{(get m k)} where \clj{m} represents the path after {\keype{\k{}}}---{\path{\keype{\makekw{b}}}{\path{\keype{\makekw{a}}}{\x{}}}}
  represents \clj{(get (get x :a) :b)}.

Figure~\ref{main:figure:hmaptyping} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
T-Assoc extends a HMap with a mandatory entry.
Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.

@@

The \classconst{} function is modeled as a constant which returns
the argument's class or \nil{}.

$$
\begin{array}{lrlr}
  \constantssyntax\\
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

@@

Our second path element \classpe{} is the symbolic representation
of a call to \classconst{}---{\path{\classpe{}}{\path{\keype{\makekw{a}}}{\x{}}}}
represents \clj{(class (get x :a))}.
Figure~\ref{main:figure:primitivesem} contains the semantics for \classconst{}.

\constantsemfigure{main}

@@

\begin{mathpar}
  {\LUpdate}
\end{mathpar}
For example, if \inpropenv {\propenv{}} {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
and
\inpropenv {\propenv{}} {\notprop {\Nil{}} {\x{}}},
then
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype{\Nil{}}} {\emptypath{}}} {\x{}}},
$$
so by the definition of \updateliteral{} (figure~\ref{main:figure:update})
\inpropenv {\propenv{}} {\isprop {\Keyword{}} {\x{}}} using \removeliteral{} (figure~\ref{main:figure:restrictremove}).

Now that we have our path elements \classpe{} and \keype{\k{}}
and the typing rules that introduce them, we give a detailed example how
each path element interacts with T-Update.

@@

\begin{minted}{clojure}
(fn [x :- (U nil Keyword)]
  (if (class x)
    ; here
    nil))
\end{minted}

@@

From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
The test has the object {\path{\classpe{}}{\x{}}} by T-App.
By T-If we can assume
{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
{\path{\classpe{}}{\x{}}}, 
by T-Update 
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
$$
This hits the sixth \classpe{} case of \updateliteral{}, which returns
$$
{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
$$
deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).

@@

\begin{minted}{clojure}
(fn [x :- '{:kw (U nil Keyword)}]
  (if (get x :kw)
    ; here
    nil))
\end{minted}

@@

$$
\rewrite {\taenv{}} {\e{}} {\ep{}}
$$
rewrites expression \e{} to a possibly-less reflective expression
\ep{} with respect to type hint environment
\taenv{} and Java class table \ct{} and
$$
\tajudgement {\taenv{}} {\hastype {\e{}} {\tatype{}}}
$$
infers the (possibly-unknown) type hint \tatype{} of expression \e{} in environment \taenv{}.

@@

We can rewrite example~\ref{example:conditionalflow} as
\inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}

@@

%Substitution is used to ensure well formed propositions.

The final line gives up---we cannot distinguish between equality and subclassing mode.

@@

To ease presentation, we present \emph{immutable}
multimethods (figure~\ref{main:figure:mmsyntax}). \defmethodliteral{} returns a new extended multimethod
without changing the original multimethod. Example~\ref{example:rep} is now written
\begin{minted}{clojure}
(let [rep (defmulti [Any -> String] class)
      rep (defmethod rep Keyword [x] (str (name x)))
      rep (defmethod rep Number [x] (str (inc x)))]
  (rep :a)) ;=> "a"
\end{minted}

@@

{\convertjavatypefigure{figure}{\footnotesize}}
